### 2016-11-29 chapter1.tex
Corrections made to IRI Section, raytracing added

\subsection{IRI Model}
\label{sec:iri}
Models can be useful tools for establishing the context of experimental measurments, especially in a global sense.  In particular, it is impractical to achieve full global coverage of all instruments, so models can be a useful way to fill in the gaps.  There are two main categories of models.  Theoretical models are based on first principles and global pictures are often found by running large computer simulations based on these first principles.  They can help give insight as to what causes specific phenomena because the process can be followed beginning to end.  Empirical models are based purely on measured data.  These can potentially be more accurate because they should replicate real measurments, but they also do not give any insight into how the system got to its ending state.  In addition, empirial models are often heavily averaged.

The international reference ionosphere (IRI) is an empirical model of the Earth's ionosphere.  It was originally created in 1969 as a joint effort between the Comittee on Space Research (COSPAR) and the International Union of Radio Science (URSI) and has been periodically updated since then \citep{Rawer1978}.  The current version is IRI-2012 \citep{Bilitza2014}, which is used throughout the work done here.  It can be accessed either as an online module or by downloading and running the FORTRAN source code, both of which are available from the International Reference Ionosphere page on the NASA website (http://iri.gsfc.nasa.gov/).

The IRI model can produce outputs of electron density, temperatures of electrons, ions, and neutrals, ion composition (O+, H+, He+, O2+, NO+, N+), and total electron content (TEC).  The model requires the input of a location (latitude, longitude, and altitude) and time (year, date, and time).  Internally, the model has the additional capability of producing profiles in any of the input parameters.  This is particularly significant for height profiles, and allows the peak heights and densities of the E and F region to be calculated.

%- Empirical model of the global ionosphere
%- Parameters that can be calculated
%- How is it available?
%	- Online
%	- FORTRAN source code
%- Biltza, 1990

>>>>

\subsection{IRI Model}
\label{sec:iri}
The International Reference Ionosphere (IRI) is an empirical model of the Earth's ionosphere.  It was originally created in 1969 as a joint effort between the Committee on Space Research (COSPAR) and the International Union of Radio Science (URSI) and has been periodically updated since then \citep{Rawer1975,Rawer1978,Rawer1981,Bilitza1985,Bilitza1986,Bilitza1990,Bilitza1997,Bilitza2001,Bilitza2008}.  The current version is IRI-2012 \citep{Bilitza2014}, which is used throughout the work done here.  It can be accessed either as an online module or by downloading and running the FORTRAN source code, both of which are available from the International Reference Ionosphere website (\url{http://iri.gsfc.nasa.gov/}).

The IRI model can produce outputs of electron density, temperatures of electrons, ions, and neutrals, ion composition (O+, H+, He+, O2+, NO+, N+), and total electron content (TEC).  The model requires the input of a location (latitude, longitude, and altitude) and time (year, date, and time).  Additionally, the model gives the option to choose between two F peak models and three bottomside thickness models.  There are a variety of other optional input parameters such as sunspot number, ionospheric index, and F10.7 radio flux which can help make the output parameters more accurate for a particular situation.  Internally, the model has the additional capability of producing profiles of any of the output parameters in either space or time.  This is particularly significant for height profiles, and allows the peak heights and densities of the E and F region to be calculated.

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make real measurements, it is important to remember that it is a model and therefor only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015}.  In particular, the height of the F-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.  Despite this, the IRI model is still a very important tool for establishing a density profile at any time in any location in the ionosphere.  And example of an altitude profile of plasma density provided by the IRI model is shown in Figure \ref{fig:densprofile}.  The displayed profiles shows how the electron density changes with altitude at local noon (day) and midnight (night) at \(75\deg\) N, \(0\deg\) E, geographic.

One of the most important uses of the IRI model has been its integration into ray tracing simulations.  It can be valuable to determine the degree to which HF radar beams refract through a dense ionosphere.  This can be done through standard raytracing tools based on numerical solutions to the Hamiltonian raypath equations \citep{Haselgrove1963,Jones1975}, however these tools require a 2D density profile along the path that the radar beam travels.  It is possible to assume a Gaussian distribution or Chapman layer as a simple model of the F-region peak, but it is much more instructive to use output from the IRI model, as it is sensitive to changes in latitude as well as seasonal and diurnal variations.
